---
title: "Regression II Final Project Group 1M"
author: "Yingzhi Ma, Guanyu Lu, Yi Yang"
date: "2023-11-26"
output:
  pdf_document: default
  html_document: default
---
Package may use in the project
```{r, message=FALSE, warning=FALSE}
library(pander)
library(ggplot2)
library(moments)
library(tidyverse)
library(psych)
library(rio)
library(MASS)
library(ResourceSelection)
library(car)
library(VGAM)
library(pROC)
library(lmtest)
```

# Upload Data from Excel File
```{r}
my_data <- read.csv("MIDUS_III_Final_Exam_Fall2023_data.csv")
#Show the first six rows of data set. 
pander(head(my_data))

my_data %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
    facet_wrap(~ key, scales = "free") +
    geom_histogram(col = 'orange')

#Show the distribution of data set. 
pander(skewness(my_data), caption = 'Skewness of numeric data')

# Check Correlation
cor <- cor(my_data);
pander(cor);
```
# Univariate Analysis
```{r}
# Continuous Variables
cont_var <- my_data %>% dplyr::select(age, cigage, alcage)
summary(cont_var)

# Categorical Variables
# List of categorical variables
cat.var.Final <- c("male", "heart", "smoke", "depress", "bp", "exercise", "health")

# Loop through each categorical variable and print its percentage frequency table
for (var in cat.var.Final) {
  cat("\nPercentage Frequency table for", var, ":\n")
  freq.table <- table(my_data[[var]])
  perc.table <- round((freq.table / sum(freq.table)) * 100, 2) # calculate percentages and round off to 2 decimal places
  pander(freq.table)
  pander(perc.table)
}
```


# Bivariate Analysis
```{r}
# Case 1
## T - Test for Continuous Predictors
t.test(my_data$heart,my_data$alcage)
t.test(my_data$heart,my_data$cigage)
t.test(my_data$heart,my_data$age)

# Calculating mean and SD for each continuous predictor within 
# each category of 'heart' for 'age'
age_stat <- my_data %>%
  group_by(heart) %>%
  summarise(
    Mean_Age = mean(age, na.rm = TRUE), 
    SD_Age = sd(age, na.rm = TRUE)
  )
print(age_stat)

# for 'cigage'
cigage_stat <- my_data %>%
  group_by(heart) %>%
  summarise(
    Mean_Cigage = mean(cigage, na.rm = TRUE), 
    SD_Cigage = sd(cigage, na.rm = TRUE)
  )
print(cigage_stat)

# for 'alcage'
alcage_stat <- my_data %>%
  group_by(heart) %>%
  summarise(
    Mean_Alcage = mean(alcage, na.rm = TRUE), 
    SD_Alcage = sd(alcage, na.rm = TRUE)
  )
print(alcage_stat)
```

```{r}
## Chi-Squared Tests for Categorical Predictors
### Loop through each categorical predictor and create a contingency table with 'heart'
for (predictor in cat.var.Final) {
  cat("\nContingency table for", predictor, "and heart:\n")
  conti.table.1 <- table(my_data[[predictor]], my_data$heart)
  perc.conti.table.1 <- prop.table(conti.table.1, margin = 2) * 100
  chi.squared.test.1 <- chisq.test(conti.table.1)
  print(chi.squared.test.1)
  print(conti.table.1)
  print(perc.conti.table.1)
}
```

```{r}
# Case 2
## Logistic Regression for Continuous Predictors
# T - Test for Continuous Predictors with 'health' as Outcome
t.test(my_data$health,my_data$alcage)
t.test(my_data$health,my_data$cigage)
t.test(my_data$health,my_data$age)

# Calculating mean and SD for each continuous predictor within each category of 'health'
# for 'age'
age_stat_2 <- my_data %>%
  group_by(health) %>%
  summarise(
    Mean_Age = mean(age, na.rm = TRUE), 
    SD_Age = sd(age, na.rm = TRUE)
  )
print("Age statistics by health categories:")
print(age_stat_2)

# for 'cigage'
cigage_stat_2 <- my_data %>%
  group_by(health) %>%
  summarise(
    Mean_Cigage = mean(cigage, na.rm = TRUE), 
    SD_Cigage = sd(cigage, na.rm = TRUE)
  )
print("Cigage statistics by health categories:")
print(cigage_stat_2)

# for 'alcage'
alcage_stat_2 <- my_data %>%
  group_by(health) %>%
  summarise(
    Mean_Alcage = mean(alcage, na.rm = TRUE), 
    SD_Alcage = sd(alcage, na.rm = TRUE)
  )
print("Alcage statistics by health categories:")
print(alcage_stat_2)
```

```{r}
## Chi-Squared Tests for Categorical Predictors
# Loop through each categorical predictor and conduct a Chi-squared test with 'health'
for (predictor in cat.var.Final) {
  cat("\nContingency table for", predictor, "and health:\n")
  conti.table.2 <- table(my_data[[predictor]], my_data$health)
  perc.conti.table.2 <- prop.table(conti.table.2, margin = 2) * 100
  chi.squared.test.2 <- chisq.test(conti.table.2)
  print(chi.squared.test.2)
  print(conti.table.2)
  print(perc.conti.table.2)
}
```


# Assumptions
```{r warning=FALSE}
# Checking for outliers in continuous variables
cont_var_long <- pivot_longer(my_data, cols = c(age, cigage, alcage), 
                                    names_to = "variable", values_to = "value") 

## Create boxplots for age, cigage, and alcage
ggplot(cont_var_long, aes(x = variable, y = value)) + 
    geom_boxplot() +
    theme_minimal() +
    ggtitle("Boxplots for Continuous Variables") +
    xlab("Variable") +
    ylab("Value")

# Checking for data consistency
summary(cont_var)

# Checking for Multicollinearity
## Correlation matrix for 'age', 'cigage', and 'alcage'
cor <- cor(my_data[c("age", "cigage", "alcage")])
print(cor)


my_data_2_3 <- my_data %>% filter(age == 64) %>% filter(bp == 1);
## Standardization for Continuous Variables in case there are multicollinearity
my_data$age <- scale(my_data$age)
my_data$cigage <- scale(my_data$cigage)
my_data$alcage <- scale(my_data$alcage)

## Case 1
### Use Linear model as a proxy 
linear_model_assump <- lm(heart ~ age + cigage + alcage + as.factor(male) + 
                              as.factor(smoke) + as.factor(depress) + 
                              as.factor(bp) + as.factor(exercise), data = my_data)
pander(vif(linear_model_assump))

## Homoscedasticity
plot(linear_model_assump, which = 1)

# Normality of Residuals
qqnorm(residuals(linear_model_assump))
qqline(residuals(linear_model_assump))

# Independence of Errors
dwtest(linear_model_assump)

## Case 2
### Use Linear model as a proxy 
linear_model_assump_2 <- lm(as.numeric(health) ~ age + cigage + 
                              alcage + as.factor(male) + as.factor(smoke) + 
                              as.factor(depress) + as.factor(bp) + 
                              as.factor(exercise), data = my_data)
pander(vif(linear_model_assump_2))

## Homoscedasticity
plot(linear_model_assump_2, which = 1)

# Normality of Residuals
qqnorm(residuals(linear_model_assump_2))
qqline(residuals(linear_model_assump_2))

# Independence of Errors
dwtest(linear_model_assump_2)
```


Case1

For the first study, you will build a predictive model for predicting if people in their midlife have ever experienced heart trouble (outcome variable). 
Step 1: select the correct model based on the distribution of the outcome variable. You might consider the following independent variables (bp, smoke, age, male, and exercise) as potential predictors. 
In step 2, you will run the model with the interaction term between smoke and male, controlling for the other variables. 
In step 3, you will assess if each model is a good fit for the data and which model (the main effects or interaction effect model) is better. Remember data cleaning and checking for potential outliers that might influence the estimates in the model is one of the major steps in statistical analysis. (NOTE: DO NOT DELETE ANY OUTLIERS, NOTE AND EVALUATE THEM).
```{r, warning=FALSE, message=FALSE}
# Step 1
my_data_1 <- my_data %>% dplyr::select(heart, bp, smoke, age, male, exercise);

#Backward and forward Model selection
backwards <- step(glm(factor(heart) ~ factor(bp) + factor(smoke) + age + 
                        factor(male) + factor(exercise), 
                     family = binomial,
                     data = my_data_1)); # Backwards selection is the default
summary(backwards);
forward <- step(backwards,
                scope = list(lower = formula(backwards),
                upper = formula(backwards)),
                direction = "forward");
formula(forward);

model_1 <- glm(factor(heart) ~ factor(bp) + factor(smoke) + age + factor(male), 
               family = binomial, data = my_data_1);
summary(model_1);
# Check Correlation for Continuous variable
vif_1 <- car::vif(model_1);
vif_1;

#Step 2
#Backward and forward Model selection
backwards_2 <- step(glm(factor(heart) ~ factor(bp) + factor(smoke) + age + 
                          factor(male) + factor(exercise) + smoke * male, 
                     family = binomial,
                     data = my_data_1)); # Backwards selection is the default
summary(backwards_2);
forward_2 <- step(backwards_2,
                scope = list(lower = formula(backwards_2),
                upper = formula(backwards_2)),
                direction = "forward");
formula(forward_2);
model_1_2 <- glm(factor(heart) ~ factor(bp) + factor(smoke) + age + 
                   factor(male) + factor(smoke) * factor(male), 
               family = binomial, data = my_data_1);
summary(model_1_2);
#Likelihood Ratio Test
lrtest(model_1, model_1_2);
```

```{r}
#Step 3
## Odds Ratios and Confidence Intervals
exp(cbind(OR = coef(model_1), confint(model_1)));
exp(cbind(OR = coef(model_1_2), confint(model_1_2)));

pander(c("Model_1 AIC" = AIC(model_1), "Model_1_2 AIC" = AIC(model_1_2), 
         "Model_1 BIC" = BIC(model_1), "Model_1_2 BIC" = BIC(model_1_2)));

## ANOVA Comparison
anova(model_1, model_1_2)

## 1- chisq probability
### Basic Model
pander(c(1 - pchisq(model_1$deviance, model_1$df.residual), 1 - pchisq(model_1_2$deviance, model_1_2$df.residual)))


## Hosmer-Lemeshow Test
### Basic Model
ht_model_1 <- hoslem.test(model_1$y, fitted(model_1), g = 10)
ht_model_1

### Interaction Model
ht_model_1_2 <- hoslem.test(model_1_2$y, fitted(model_1_2), g = 10)
ht_model_1_2

# Plot ROC curves and AUC
rocplot1 <- roc(heart~fitted(model_1), data=my_data_1);
plot.roc(rocplot1, legacy.axes=TRUE);
rocplot2 <- roc(heart~fitted(model_1_2), data=my_data_1);
plot.roc(rocplot2, legacy.axes=TRUE);
pander(c("Model_1 AUC" = auc(my_data_1$heart, model_1$fitted.values), 
         "Model_1_2 AUC" = auc(my_data_1$heart, model_1_2$fitted.values)));
```

Case2
For the second case study, build a predictive model using a self-rated health category for individuals in their midlife as an outcome variable. Select the correct model based on the distribution of the outcome variable. 
In step 1, you will fit and conduct the main effects model with the variables depress, alcage, cigage, age, and bp. 
In step 2, you will test any relevant assumptions. 
In step 3 [2 bonus points], predict probabilities for those with diagnosed high blood pressure if (1) age = 64 and depress = 0, and (2) if age = 64 and depress = 1. Report the difference in probabilities per group (i.e., excellent, good, fair, poor).
```{r,warning=FALSE}
# this is a cumulative logit model for ordinal response. 
my_data_2 <- my_data %>% dplyr::select(health, depress, alcage, cigage, age, bp);
#Step 1
model_2 <- vglm(factor(health) ~ factor(depress) + alcage + cigage + age + factor(bp), 
                family = cumulative(parallel=TRUE), data = my_data_2);
summary(model_2);

vif_model <- lm(health ~ depress + alcage + cigage + age + bp, data = my_data_2);
vif_2 <- vif(vif_model);
vif_2;
#Step 2
## Relative Odd Ratio
(fit.rr.2.1 <- exp(coefficients(model_2)))

## Consider only the significant variables in the model
table(my_data_2$health, my_data_2$depress)
table(my_data_2$health, my_data_2$bp)

#For depress
odd.ratio.dp1 <- ((77+212+619)*50)/((23+45+40)*913)
odd.ratio.dp2 <- ((77+212)*(40+50))/((23+45)*(619+913))
odd.ratio.dp3 <- (77*(50+40+45))/(23*(212+619+913))
odd.ratio.dp <- odd.ratio.dp1 + odd.ratio.dp2 + odd.ratio.dp3

#For bp
odd.ratio.bp1 <- ((19+85+281)*382)/((81+172+378)*581)
odd.ratio.bp2 <- ((19+85)*(382+378))/((81+172)*(581+281))
odd.ratio.bp3 <- (19*(382+378+172))/(81*(581+281+85))
odd.ratio.bp <- odd.ratio.bp1 + odd.ratio.bp2 + odd.ratio.bp3

odd.ratio.case2 <- data.frame(
  'Variable' = c('odd.ratio.dp1', 'odd.ratio.dp2', 'odd.ratio.dp3', 'odd.ratio.dp',  
                 'odd.ratio.bp1', 'odd.ratio.bp2', 'odd.ratio.bp3', 'odd.ratio.bp'),
  'Value' = c(odd.ratio.dp1, odd.ratio.dp2, odd.ratio.dp3, odd.ratio.dp, 
              odd.ratio.bp1, odd.ratio.bp2, odd.ratio.bp3, odd.ratio.bp)
)
pander(odd.ratio.case2)

#Step 3
#my_data_2_3 <- data.frame(depress = c(0, 1), alcage = mean(my_data$alcage, na.rm = TRUE), 
                      # cigage = mean(my_data$cigage, na.rm = TRUE), age = 64, bp = 1)
result <- predict(model_2, newdata = my_data_2_3, type = "response");
not_equal_to_1_indices <- matrix(nrow = nrow(result))
for (i in 1:nrow(result)) {
  if (sum(result[i,]) != 1) {
    not_equal_to_1_indices[i,] <- i
  }
}
not_equal_to_1_indices
```